Asymmetric relationships and their effects on coexistence: Supporting information

### 2023-08-01 paper/RefereeResponses1.tex

\begin{document}

\noindent \emph{Dear Editors,} \\

\noindent \emph{Thank you for the useful and very positive reviews of our manuscript, and for
the opportunity to resubmit a revised version. We have addressed all referee comments below and in
the manuscript, and hope the paper can now be rapidly accepted. } \\

\noindent \emph{We have replied to referee comments below in italic, with each response preceded
by ``***'', to make our responses easier to find. We submit the following pdf files: The main text pdf, including all figures; 
the Supporting Information; and a version of the main text with changes tracked. The document showing 
the changes we made does not include links such as citations and
references to figures, but it does accurately represent the changes made to the text.} \\

@@

My first comment related to this is that some form of this text (whcih is relatively short---around half a page in the SI) and, importantly, the Figure should be in the main text.  I think this is such a key part of the definitions that are being used for the rest of the paper that readers will want to get an intuition for what the "partial sharp" distributions look like and how they are arrived at. \\

\noindent ***\emph{We have done what the referee requests, here, and we agree with this request. Thanks for
making the request. This request is somewhat in tension with other requests, from other referees, 
to minimize the quantitative 
content of the main text, and to keep the overall length under the journal limit while adding a bunch of
verbal explanations requested by the other referees. We point out this tension here, and refer back to it in
other responses, below. We also point out that we have not only moved the figure back to the main text,
we have also added a panel to the figure in response to the related comment C13, below.} \\
%DONE
%DAN: In other replies, you can refer to this reply to justify when you
%decide to add explanations to the SI instead of to the main text, and when you have to respond to requests
%to minimize the quant.

\noindent C5) Perhaps equally important to clarifying this process for the readership is whether this process is unique. I could not quite get a sense of whether there could be alternative versions of this removal of ATA from a bivariate distribution, that still satisfy the authors' constraints on marginal distributions and overall correlation between the two variables. Are there alternatives, or is the authors' procedure unique in achieving the desired outcome?  If the former, then I would want to understand if the outcomes downstream tend to be robust to the specific method used. If the latter, it would be helpful to have an explanation as to why this is a unique decomposition. \\

\noindent ***\emph{This is a good question, thanks for asking it. The extent to which the process is 
unique is now covered in a new SI section, refered to from the main text at the end of section 2.3, 
near the end of the text the referee asked us to move to the main text. There is a choice involved,
but it is a pretty canonical choice -- certainly more canonical than the kinds of choices which are often made
in MCT, e.g., $E$ and $C$ can often be defined in more than one way. Basically the choice involves 
using a bivariate normal distribution instead of another, more esoteric 
distribution with standard normal marginals and symmetric tail association. The bivariate normal is by far the 
most natural example of such a distribution. See the new SI section for further explanations. }

\emph{We are reticent, at this time, to launch into a lot of robustness explorations with regard to this choice, 
as the referee may be suggesting, for several reasons. First, other referees and the editor have urged us to
simplify, and such explorations would complexify. Second, as mentioned above, the choice we made is very
natural and standard, so exploring other possible choices seems of 
secondary importance. Finally, our overall goal in writing this paper
was to show that ATAs may be important for coexistence and, as such, deserve additional exploration in this
context by the wider research community. As long as any choices we made are reasonable (they are, very much so), our point is
made, regardless of whether making some esoteric choice might have altered outcomes. As we already argue in the
Discussion, since we have now showed that ATAs \textbf{can} be important, future work should explore \textbf{when} 
they are important, and part of 
that exploration could be whether some more esoteric choice for $(e_i c_{i \bs i})$ might alter our results. }\\
%DONE

@@

\noindent C6) The notation by which the partial sharp variables are introduced was a little unintuitive to me.  For example, when the authors compare the expectation values 
$E[r(E,C)]$ and $E[r(E^{||},C^{||})]$, what is really changing is the distribution over which the expectation values are evaluated.  Maybe this is standard notation, but somehow it may be clearer to put the subscript on the E for expectation value, to indicate that the probability distribution used to generate the definition of the averaging is what differs between the two expressions, while the variables E and C themselves are in the end just dummy variables to be integrated over.  \\

\noindent ***\emph{This comment has led us to put some clarifying technical comments in a new SI section 2,
so thanks for making the comment.}

\emph{We use notational conventions of random variables which are formally correct and standard, 
as we now explain in more detail in the SI. The alternatives the referee seems to suggest would be misleading to certain audiences 
because they would be somewhat in conflict with existing conventions of probabilistic notation.
$E_i$, $C_{i \bs i}$, $E_i^{\nat}$, $C_{i \bs i}^\nat$ and the earlier 
notation $E_i^\shp$ and $C_{i \bs i}^\shp$ all denote \textbf{random variables}, using a 
notational approach which is standard in applications of probability. 
Whereas the \textbf{distributions} of the random variables $E_i$, $E_i^{\nat}$ and $E_i^\shp$ (for instance) are all the same,
these things are \textbf{not} the same \textbf{as random variables} because their relationships with other random variables 
differ. The relationships between random variables are considered instrinsic
to the notation. Thus it's not that somehow the expected value in $\E[r(E,C)]$ is a different definition of expected
value than the one in $\E[r(E^{||},C^{||})]$ -- there is only one expected value operator, and we dare not 
introduce different versions because we do not want to confuse all those who already know the commonly
accepted definition of expected value. 
Instead, $r(E,C)$, $r(E^{||},C^{||})$ and $r(E^\shp,C^\shp)$ differ from each other 
as random variables because of the distinct relationships between their
different $E$ and $C$ constituents. It would be formally 
incorrect, and confusing to readers familiar with probability, to try to define multiple expected value 
operators, when what fundamentally differs, here, is the bivariate random variables 
themselves. Readers familiar with
probability already know what the one accepted definition of expected value is, and would typically
not be comfortable with providing additional definitions. Some concepts from probability
which illustrate these conventions are as follows. Recall that if $X$ and $Y$ are random variables, the expectation of the 
product, $\E(XY)$, depends not only on the univariate distributions of $X$ and $Y$, but also on the relationship
between $X$ and $Y$, namely their joint distribution. In particular, $\E(XY)=\E(X)\E(Y)$ if $X$ and 
$Y$ are independent, but not necessarily otherwise. These definitons and notation are 
standard in probability theory; notice that we use the same definition of $\E$ regardless of 
whether $X$ and $Y$ are independent or nonindependent, it's just that the product random variable
$XY$ differs accordingly so you get different results of $\E(XY)$ in those distinct cases.
One should not attempt to instead view the difference between $\E(XY)$ for $X$ and $Y$ independent
as distinct from $\E(XY)$ for $X$ and $Y$ non-independent by looking at two different versions
of expected value and considering $X$ and $Y$ to be ``dummy variables'', somehow.}

\emph{Our notation also parallels that of Ellner et al (2016, 2019). To make it
easier for readers to understand and to reduce the potential for future confusion, it makes sense to continue
using existing notation whenever that notation is formally correct and otherwise effective (Ellner's notation is 
both of these, as explained above).}

\emph{Nevertheless, the referee's comment has made us understand that some readers unfamiliar with random variable 
notation may benefit from putting these explanations in the SI, so we have done so.} \\
%DONE

@@

\noindent C7) It was not clear to me what the reasoning was for including the correlation and ATAs in abundance for the two planktonic species in Figure 1(d) and (e).  These ATAs in abundances don't in an obvious (to me) way translate into ATAs of the underlying environment and competition variables for each species. Is this really a demonstration that ATAs occur for C and E?  If not, then it seems to be like saying ATAs exist somewhere in nature, and therefore might be present in our focal situation, which is not such a very strong reason to show (d) and (e).  Maybe I am missing something. \\

\noindent ***\emph{Yes, the referee is right about this, we are not illustrating ATAs in $E$ and $C$ at this stage of the 
paper - the reader would anyway not be equiped to handle that at this stage, since we have not even formally defined 
$E$ and $C$ at the point this figure is first cited (in the very first paragraph). 
We were asked to include this kind of empirical illustration 
by poeple who read the manuscript before we submitted it, so it seems important to some readers to have this. 
The value is not only in showing that ATAs occur in nature, but is primarily  
in helping people understand more fully what ATAs look like when they occur, by giving a real example. 
We appreciate what the referee is saying,
here, that perhaps one non-$EC$ example of ATAs from nature is not going
to give much motivation for looking for ATAs in $E$ and $C$. However, we have published droves of other examples of
ATAs in nature, and we now cite these (changing what used to be ``Fig. 1d,e show contrasting examples of ATAs in nature
using plankton population density time series'' to ``Fig. 1d,e show contrasting examples of ATAs in nature
using plankton population density time series; many other examples of ATAs in nature have also been explored (Ghosh et al 2020)''). 
And anyway we cannot produce a prior example of ATAs between $E$ and $C$ at this early point in the manuscript, 
as the referee may be suggesting, because there are no prior examples - this paper is, to our
knowledge, the first time anyone has looked at ATAs between $E$ and $C$. Fortunately, we think the other motivational
content of the Introduction is strong, and should stand on its own to motivate readers to spend their mental 
energy reading this paper. This motivational content has also been strengthened since the prior version - see other replies. 
In partcular, we show in the Introduction, for one simple simulation, that ATAs between E and
C make a visually striking difference between simulation outcomes that appears clearly to influence coexistence.
We also provide an intuitive example with water limitation of why one might expect ATAs between $E$ and $C$, 
a bit later in the Introduction from the point to which the referee's comment pertains.
And we also provide a new paragraph explaining, intuitively, how ATAs work. } \\
%DONE

@@

\noindent C8) The authors mention that $r_{i \bs i}$ indicates growth rate of species when $i$ is rare and $j$ is at steady state.  I think $r_{j \bs i}$ must also assume that $j$ is at steady state, but this isn't mentioned. Perhaps that could be added for completeness. \\

\noindent ***\emph{This was done already, the referee appears to have missed it. It was line 136 in the original submission,
now line 147 in the resubmitted version.} \\
%Don't forget to check line numbers at the end.

@@

\noindent ***\emph{We thank the referee for this viewpoint. We also thank the referee for acknowledging that this
is their opinion, and that data are insufficient for a clear conclusion to be reached at this point. We find that 
such an acknowledgment is both rare from a referee and very useful for causing us, as authors, to most usefully revise our 
paper. We have added new discussion material covering these topics and find that the new material improves the paper.
We tried to robustly represent the referee's viewpoints in the new material, and we cited the papers the referee mentions.}

\emph{We point out that we worked not only
with the lottery model, but also with an empirical diatom system. That system definitely supported the importance of 
ATAs for coexistance, which would tend to give at least one data point in favor of the importance of these phenomena, though we 
acknowledge that our lab systems is also a system ``where niche differentiation via ... predators is impossible.''
It may also be a system ``where niche differentiation via different resources ... is impossible,'' though resource 
differentiation can happen when you don't expect it. It seems to us 
the first step in assessing the importance of a newly recognized phenomenon has to be examining it in simple systems
such as we have used, prior to examing it in field systems, in all their complexity, where niche differentiation
by resources and predators can more readily occur. But once our paper is published we hope and expect that we and
others will more broadly assess the potential importance of ATAs in such ``in the wild'' systems.}

@@

\emph{We also point
out (and the referee already seems to accept this) that it is possible, \emph{a priori}, for ATAs to be 
important for coexistence in a system even while storage effects are not.
Storage effects are ATA effects plus what we called, in the manuscript, the effects of $EC$ correlation \emph{per se} 
($\Delta_i^{[E \nat C]}$ in our notation). Either of these two summands can be negative or positive.
So if the two effects are opposite in sign and similar in magnitude, storage effects will be unimportant for 
coexistence, but ATA effects will be important.}

\emph{So the referee appears to agree with us that storage effects and ATA effects may or may not turn
out to be broadly important in ecology. Our research goals were to introduce the idea of ATA effects on
coexistence and to show that ATAs \textbf{may} be broadly important for coexistence, by showing examples where they
\textbf{are} important. The few examples we showed are all that are readily available at this time.
We won't know for sure if ATAs (or storage effects) \textbf{are} important until more data can be brought to bear. 
Future work can assess, broadly, how important ATAs are by looking across a larger number of systems, once
available. These are two different goals. It seems crucial to us not to allow insufficiently tested 
opinions about the second set of goals to obstruct the publication of results pertinent to the first set of 
goals, since doing so would keep a lid on the basic ideas and precent their later broad testing.
So we thank the referee for not doing that. We modified a few key points in the text to clarify 
our viewpoint that we do not think our work definitively indicates that ATAs will certainly be a crucial
mechanism of coexistence; instead we just think our work opens that possibility, for future exploration.
For instance, we replaced the last sentence of the first paragraph of the Discussion, which used to be
``Though future work should seek to understand precisely when ATAs are or are not important
for coexistence, our work demonstrates the overall importance of this new mechanism'' with the revision
``Future work should seek to understand precisely when ATAs are or are not important
for coexistence. Our work demonstrates the potential for ATAs to be an important mechanism of coexistence, generally.''
And we changed the last sentence of the abstract, which used to read ``ATA influences are an important new 
mechanism of coexistence'' so that it now reads ``ATA influences may be an important new mechanism of coexistence.''
We think these changes make the paper stronger, so thanks for spurring us to make them. The last
sentence of the Introduction already read ``Overall, our study presents a new mechanism of species coexistence
and a means of understanding its theoretical and empirical importance,'' which accurately reflects our belief
that what we've accomplished is an indication that ATAs \textbf{may} be important for coexistence, 
and a means of assessing
that in future work, rather than a definitive conclusion that they are important for coexistence.}

\emph{The new Discussion material is in an SI section which is cited from the main text Discussion section. 
We would have liked to put some of the new material into the main text. But length limits, together with
clear requests from other referees to add a variety of explanatory content to the main text, made doing so impossible.} \\
%DONE

@@

\noindent C16) L41: The “we” is slightly confusing because the author teams share only one author \\

\noindent ***\emph{We changed the wording here.} \\
%DONE

\noindent C17) L48: I do not know these on the top of my head, and likely many other readers will neither, so maybe quickly refresh our memory \\

\noindent ***\emph{There is no space to intro these things, so we limited this text to just mentioning 
extinction risks and stability of ecosystem functioning. Whereas some ecologists may not have heard of 
Taylor's law, we feel confident that extinction risks and stability of ecosystem functioning will ring
a bell for most ecologists. It was not actually necessary to link to Taylor's law, we just thought it 
would provide an additional tie-in to another part of the literature. But that won't work if the referee is right 
that most ecologists will become confused, so we were happy to drop that connection here, leaving this
section of the text to stand on the connections to extinction risk and ecosystem stability.} \\
%DONE

@@

\noindent C18) L49: This is not immediately clear to me why (obviously, you’ve shown this in the other paper). Is it possible to briefly state why? \\

\noindent ***\emph{It is not necessary for the reader to understand the mechanism we explored in that other
paper in order to understand this current paper, we were just trying to tie the current work in to 
another area of research in ecology, for the sake of contextualizing it a bit and motivating studies of ATAs generally. 
So an explanation, here, is not 
strictly necessary to the current paper. Nevertheless, we tried to satisfy the referee's request of adding a short bit of explanation
without disrupting the flow too much. Thanks for your interest, by the way, we encourage you to have a look at the
other paper if you haven't already.} \\
%DONE

@@

\noindent C19) L61: What if not? Or many readers have only rough understanding of it, maybe quickly refresh or refer to later \\

\noindent ***\emph{It is not necessary for the reader to understand what these are at this point, and,
in fact, the reader does not need to understand what relative nonlinearity is at all for the main text
and main ideas of the paper. We apologize for giving the false impression, here, that more background knowledge was needed
than is actually the case. We eliminated the reference to relative nonlinearity and indicated to the 
reader that storage effects would be conceptually reviewed below (in the very next paragraph) and then 
defined in Theory, which they are.} \\
%DONE

@@

\noindent C20) Sidenote: After writing these few comments it appears to me that you seem to have a very knowledgeable reader in mind, which may not be the case for a general journal as Ecology letters. \\

\noindent ***\emph{Well, we appreciate the referee's effort to spur us to produce the clearest 
paper possible, and we have used their comments to clarify the paper. But, of the comments 
C16-C19 above, only C19 relates at all to whether this paper is written for a general ecological audience,
and we have revised the text C19 mentions so as to robustly address that comment (see response above). 
The manuscript text mentioned in C17-C18 was just linking this paper to other prior work on ATAs. 
It's not necessary for the reader to understand that work to understand this paper, the point is just to make 
the link to that other literature, so that readers can explore it if they so choose, 
and to show that ATAs have been found to be important in ecology, generally, to motivate the study.
Furthermore, we don't think our paper is any more specialized or mathematical than the papers Ellner et al (2016, 2019) which
we cite, and which inspired our paper, and which were published in Ecology Letters. Nor is it more 
specialized than many other papers we have read in Ecology Letters. Quantitive approaches are absolutely 
central in ecology these days, and increasingly so. So comments which would tend to sideline such papers to
"specialist journals", which fewer people read, seem outdated to us, especially when the papers in question 
are about mainstream topics such
as coexistence. Nevertheless, we do robustly agree that every individual paper should be made as accessible
as possible, given the subject matter, and we have gone through the paper and redoubled our efforts to make the 
paper broadly accessible.} \\
%DONE

%DAN: made it to here 2023 08 01, 609am

### 2023-08-01 paper/SupMat.Rnw
pretty much done

Given scatterplot data, $(x_i,y_i)$ for $i=1,\ldots,N$, \cite{Ghosh_2020_AER} defined the 
\emph{normalized rank} $u_i$ of $x_i$ as the rank of $x_i$ in the set 
$\{x_1,\ldots,x_N\}$ divided by $N+1$, where the smallest value is taken to have rank $1$. 
The normalized ranks $v_i$ were defined for $y_i$ in a likewise manner. 
Then, the \emph{partial
Spearman correlation} for the left halves of the distributions, $\cor_{\text{left}}$, was defined as 
\begin{equation}
\frac{\sum(u_i-\mean(u))(v_i-\mean{v})}{(N-1)\sqrt{\var(u)\var(v)}},
\end{equation}
where the sum is over all indices $i$ for which $u_i+v_i<1$ (though the means and variances
in the above formula are computed over all the data). The indices satisfying $u_i+v_i<1$
are those corresponding to points falling under the dashed diagonal lines in Fig. \ref{MT-fig:pedagog1}d-e.
%***DAN: Note I mentioned a feature which Jasmin has not yet added, so make sure she 
%adds it to those panels.
The partial Spearman correlation for the right halves of the distributions,
$\cor_{\text{right}}$, was defined using the same formula but where the sum is 
now over all indices $i$ for which $u_i+v_i>1$. These are the points falling above the
dashed diagonal lines in Fig. \ref{MT-fig:pedagog1}d-e. The left and right partial Spearman
correlations typically add up to the Spearman correlation. The index 
$\cor_{\text{left}}-\cor_{\text{right}}$ is the index used by \cite{Ghosh_2020_AER}
mentioned in the previous paragraph.

@@

To generate $M$ points $\left(b_1^{(i)}, b_2^{(i)}\right)$, $i=1,..,M$, for the left-tail association 
case (Fig. \ref{MT-fig:pedagog1}a), we first generated M points $(a_1^{(i)},a_2^{(i)})$, $i=1,...,M$, from a bivariate normal 
distribution $N\left( \vec{0}, \Sigma \right)$, where $\Sigma = \begin{pmatrix} 1&0\\0&1\end{pmatrix}$. 
Then, for each index $i$, we randomly, with 50\% probability, either 
(A) let $b_1^{(i)} = -|a_1^{(i)}|$ and $b_2^{(i)} = -|a_1^{(i)}|$, or 
(B) let $b_1^{(i)} = |a_1^{(i)}|$ and $b_2^{(i)} = |a_2^{(i)}|$. 
Perfect association in the left tails of the resulting distribution $\left(b_1,b_2\right)$ results from 
the fact that, in case (A) above, both $b_1^{(i)}$ and $b_2^{(i)}$ were assigned to the same value. 
Right-tail associated noise (Fig. \ref{MT-fig:pedagog1}c) was generated in an analogous way.

@@

Fig. \ref{fig:lognormal_fecundidities_explanatory} is similar to 
Fig. \ref{fig:beta_fecundidities_explanatory}, but is based on the log-normal 
fecundities lottery model instead of the beta fecundities lottery model. 
The figure illustrates that the potential for storage events is equivalent in
the upper and lower tails of the $EC$ distribution (labelled ``right'' and ``left'',
respectively, on panels A-C). That is why ATA effects are modest for the 
log-normal fecundities lottery model for the both the left- and right-tail
associated cases, and are negligible for the example parameters considered in 
Fig. \ref{fig:lognormal_fecundidities_explanatory}.

>>>>

Given scatterplot data, $(x_i,y_i)$ for $i=1,\ldots,N$, \cite{Ghosh_2020_AER} defined the 
\emph{normalized rank} $u_i$ of $x_i$ as the rank of $x_i$ in the set 
$\{x_1,\ldots,x_N\}$ divided by $N+1$, where the smallest value is taken to have rank $1$. 
The normalized ranks $v_i$ were defined for $y_i$ in a likewise manner. 
Then, the \emph{partial
Spearman correlation} for the left halves of the distributions, $\cor_{\text{left}}$, was defined as 
\begin{equation}
\frac{\sum(u_i-\mean(u))(v_i-\mean{v})}{(N-1)\sqrt{\var(u)\var(v)}},
\end{equation}
where the sum is over all indices $i$ for which $u_i+v_i<1$ (though the means and variances
in the above formula are computed over all the data). The indices satisfying $u_i+v_i<1$
are those plotted using open points in Fig. \ref{MT-fig:pedagog1}d-e.
The partial Spearman correlation for the right halves of the distributions,
$\cor_{\text{right}}$, was defined using the same formula but where the sum is 
now over all indices $i$ for which $u_i+v_i>1$. These are the filled points 
in Fig. \ref{MT-fig:pedagog1}d-e. The left and right partial Spearman
correlations typically add up to the Spearman correlation. The index 
$\cor_{\text{left}}-\cor_{\text{right}}$ is the index used by \cite{Ghosh_2020_AER}
mentioned in the previous paragraph.

@@

To generate $M$ points $\left(b_1^{(i)}, b_2^{(i)}\right)$, $i=1,..,M$, for the left-tail association 
case (Fig. \ref{MT-fig:pedagog1}a), we first generated M points $(a_1^{(i)},a_2^{(i)})$, $i=1,...,M$, from a bivariate normal 
distribution $N\left( \vec{0}, \Sigma \right)$, where $\Sigma = \begin{pmatrix} 1&0\\0&1\end{pmatrix}$. 
Then, for each index $i$, we randomly, with 50\% probability, either 
(A) let $b_1^{(i)} = -|a_1^{(i)}|$ and $b_2^{(i)} = -|a_1^{(i)}|$, or 
(B) let $b_1^{(i)} = |a_1^{(i)}|$ and $b_2^{(i)} = |a_2^{(i)}|$. 
Perfect association in the left tails of the resulting distribution $\left(b_1,b_2\right)$ results from 
the fact that, in case (A) above, both $b_1^{(i)}$ and $b_2^{(i)}$ were assigned to the same value. 
Right-tail associated noise, $\left(b_1^{(i)}, b_2^{(i)}\right)$, $i=1,..,M$ (Fig. \ref{MT-fig:pedagog1}c) 
was generated by taking the negative of left-tail noise.

@@

Fig. \ref{fig:lognormal_fecundidities_explanatory} is similar to 
Fig. \ref{fig:beta_fecundidities_explanatory}, but is based on the log-normal 
fecundities lottery model instead of the beta fecundities lottery model. 
The figure illustrates that the potential for storage events is equivalent in
the upper and lower tails of the $EC$ distribution (labelled ``right'' and ``left'',
respectively, on panels A-C). That is why ATA effects are modest for the 
log-normal fecundities lottery model for the both the left- and right-tail
associated cases, and are negligible for the example parameters considered in 
Fig. \ref{fig:lognormal_fecundidities_explanatory}. A careful reader may notice the
upper horizontal-axis histograms on Fig. \ref{fig:lognormal_fecundidities_explanatory}D, F
are mirror images of each other. This was because the right-tail noise with standard normal
marginals that we used was generated
by simply taking the negative of our left-tail noise with standard-normal marginals (SI section 
\ref{sect:noise}), a choice that could not realistically have influenced results in a meaningful way.
This can also be seen by comparing main text Fig. \ref{MT-fig:pedagog1}a, c.
